Efficiency

### 2018-06-05 DraftExec.tex
latest, with typos for senate

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs that might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Exactly when and how are pumpjets quieter than propellers?]
It's generally accepted that the most important acoustic consideration is preventing cavitation from occurring. Whilst all turbulence and other disturbances underwater create some noise, becuase water is incompressible, quite a lot of `round and round' movement (as in turbulence) still results in relatively little `in and out' movement, which is what tends to result in pressure waves propagating away from the sources.  When gaps open up in the water and subsequently collapse, as is the case in cavitation, the noise produced is much louder.
\\
\\
Pumpjets have an acoustic advantage because they can avoid cavitation in circumstances where an open propeller might begin to cavitate, in particular at higher speeds. However, at lower speeds such as patrol speeds (and also at greater depths) propellers can easily avoid cavitation because they don't need turn very fast in the water. Consequently, the avoidance of cavitation in these circumstances offers no advantage.  In addition, at very low speeds pumpjets experience increasing levels of flow separation, which creates unsteady vortices before and between the blades of the rotor.  This degrades the acoustic performance of the jet, particularly at low frequencies, which are harder to shield, and tend to carry much further. At some very low speed, a well-designed propeller will necessarily be much quieter than a jet, though both may still be much quieter than if cavitation occurred.
\\
\\
This means that the selection of a propulsion system for acoustic advantage is highly dependent on the probable speed profiles of submarine's operations. For a nuclear submarine, which can operate at higher speeds (above 15 or 20kt) for as long as they like, the ability to remain relatively quiet at these speeds is extremely desirable. It would allow them to undertake fast transits, or stalk enemy shipping, while remaining very quiet.  Diesel electric submarines simply cannot sustain these sorts of speeds for any more than a few hours, and hence they are unlikely to use them unless in an emergency or critical tactical manoeuvre.  Almost all of their sensitive or combative operations are likely to be at very low speeds, and hence being as quiet as possible in absolute terms at low speeds is most likely to be advantageous.  The selection of a pumpjet on a diesel-electric submarine on acoustic grounds is consequently a strange choice, as it amounts to elevating a tactical scenario which demands a short, high-speed burst above all other circumstances where a long, silent dive at low speed is required.
\end{tcolorbox}

>>>>

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst the considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one was confident enough on the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found this seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs that might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance and consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Exactly when and how are pumpjets quieter than propellers?]
It's generally accepted that the most important acoustic consideration is preventing cavitation from occurring. Whilst all turbulence and other disturbances underwater create some noise, becuase water is incompressible, quite a lot of `round and round' movement (as in turbulence) still results in relatively little `in and out' movement, which is what tends to result in pressure waves propagating away from the sources.  When gaps open up in the water and subsequently collapse, as is the case in cavitation, the noise produced is much louder.
\\
\\
Pumpjets have an acoustic advantage because they can avoid cavitation in circumstances where an open propeller might begin to cavitate, in particular at higher speeds. However, at lower speeds such as patrol speeds (and also at greater depths) propellers can easily avoid cavitation because they don't need to turn very fast in the water. Consequently, the avoidance of cavitation in these circumstances offers no advantage.  In addition, at very low speeds pumpjets experience increasing levels of flow separation, which creates unsteady vortices before and between the blades of the rotor.  This degrades the acoustic performance of the jet, particularly at low frequencies, which are harder to shield, and tend to carry much further. At some very low speed, a well-designed propeller will necessarily be much quieter than a jet, though both may still be much quieter than if cavitation occurred.
\\
\\
This means that the selection of a propulsion system for acoustic advantage is highly dependent on the probable speed profiles of submarine's operations. For a nuclear submarine, which can operate at higher speeds (above 15 or 20kt) for as long as they like, the ability to remain relatively quiet at these speeds is extremely desirable. It would allow them to undertake fast transits, or stalk enemy shipping, while remaining very quiet.  Diesel electric submarines simply cannot sustain these sorts of speeds for any more than a few hours, and hence they are unlikely to use them unless in an emergency or critical tactical manoeuvre.  Almost all of their sensitive or combative operations are likely to be at very low speeds, and hence being as quiet as possible in absolute terms at low speeds is most likely to be advantageous.  The selection of a pumpjet on a diesel-electric submarine on acoustic grounds is consequently a strange choice, as it amounts to elevating a tactical scenario which demands a short, high-speed burst above all other circumstances where a long, silent dive at low speed is required.
\end{tcolorbox}
